Draft_Application_Stochastic_Volatility

### 2021-07-15 Real Dataset SV.Rmd
Some changes

output:
  pdf_document: 
   latex_engine: xelatex
   fig_width: 5
   fig_height: 3.5
   fig_caption: true
   extra_dependencies: ["float"]
   number_sections: true

---

\section{SV models + Literary Review}
In section (\textit{fill with the number of the section}) a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset. Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows,
\begin{align*}
    y_t|x_t & \sim N(0,e^{x_t})\\
    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align*}

@@

Stochastic Volatility Models, whose early formulation is commonly attributed to Taylor (1982, 1986), allow to account for time-varying and autoregressive volatility in financial returns, posing themselves as a valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models in dealing with non-constant volatility. Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as
\begin{align*}

@@

\end{align*}
where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$. Kim, Shephard and Chib (1998) prefer $\psi=1$.} Most importantly, $\epsilon_t$ and $\eta_t$, the Gaussian white noise processes that drive the canonical model, are assumed to be uncorrelated. This latter assumption justifies another definition of this specification, namely "discrete SV model without leverage".\footnote{Note that in this review we do not focus on continuous time SV models. In fact, such models have attracted a considerable amount of research in financial econometrics and mathematical finance, especially after Hull and White (1987) considered stochastic volatility for option pricing. Arguably, the most influential model was then proposed for option pricing by Heston (1993), a SV model with leverage effects and square root diffusion driving volatility. Diffusion-based SV models enjoyed increasing popularity, see for example Barndorff-Nielsen and Shephard (2001), or Christoffersen et al. (2010), who investigate alternatives to the entrenched affine square root SV model.  Eraker (2004) proposed a SV model with correlated jumps in prices and volatility, extending Heston's model, while Comte and Renault (1998) extended Hull and White's model as to feature long memory properties. More recently, Gatheral et al. (2018) further built on Comte and Renault's fractional SV model to propose the popular "rough volatility models" (see also Friz et al., 2021).}\
In order to accommodate for leverage effects, namely capture the increase in volatility that follows a drop in the returns, the model can extended as in Omori et al. (2007),
\begin{align*}

@@

\end{align*}
where $\rho<0$ captures the negative correlation. Such specification, which can be referred to as "discrete SV model with leverage", captures the asymmetric response of volatility to returns of different signs, so that similar specifications are sometimes also deemed as "asymmetric SV models" (e.g. Harvey and Shephard 1996, Mao et al. 2020). \

@@

Assuming that $\epsilon_t$ follows a Student-$t$ distribution, and exploiting the fact that $\epsilon_t$ can then be written as $\lambda_t^{-1/2}\zeta_t$, where $\zeta_t \sim N(0,1)$ and $v\lambda_t \sim \chi^2_v$ (Harvey et al. 1994, Chib et al. 2002),  we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004),
\begin{align*}

@@

\end{align*}
To capture other elements of the behaviour of financial data, several other extensions have been proposed. For example, SV models have been extended to include conditional heteroskedasticity in the mean returns (Koopman and Hol Uspensky 2002) to capture potential volatility feedback effects, or to feature autoregressive moving average innovations (Chan 2013, Zhang et al. 2020), allowing better goodness of fit and out-of-sample forecasts.\footnote{Dimitrakopoulos and Kolossiatis (2020) note that "the moving average component, the leverage effect and the conditional heteroscedasticity in mean have been considered separately in the stochastic volatility literature" and provide two specifications, one featuring an MA component and leverage effects, the other an MA component and conditional heteroskedasticity in mean.} \footnote{Other extensions, though ones for which we do not report the specifications, feature modelling the latent volatility process $x_t$ as an ARFIMA process (Long Memory Stochastic Volatility model, Breidt et al. 1998) or as governed by a first-order Markov process (Markov Switching Stochastic Volatility model, So et al. 1998). Recently, Luo et al. (2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as
\begin{align*}

@@

\end{align*}
while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads
\begin{align*}

@@

\end{align*}
where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\

@@

Finally, before moving to the application of Sequential Monte Carlo techniques to a SV model, we conclude by going through some references for the Bayesian analysis proposed for such models. Starting from the seminal work of Jacquier et al. (1994), the use of MCMC methods has become increasingly popular for parameter estimation and smoothing exercises in SV models (e.g. Kastner 2019, presenting the R package \textit{stochvol} for Bayesian parameter estimation, and Chopin and Papaspiliopoulos 2020, who use MCMC to sample from the smoothing distribution of a SV model). As regards filtering exercises, the adoption of particle filters was rather rapid:\footnote{Actually, SV models are now often used as straightforward applications of particle filters on non-linear state space models, see for example Andrieu et al. (2010), Douc et al. (2014) or Chopin and Papaspiliopoulos (2020).} indeed, latent volatilities in Kim et al. (1998) were already filtered by employing the particle filter suggested in Pitt and Shephard (1999), paving the way for subsequent applications (\textit{inter alia}) in Chib et al. (2006), Omori et al. (2007), Kim and Stoffer (2008) and 
Nakajima and Omori (2012).\

@@

As previously mentioned, we will analyze the behaviour of the described filtering tools associated to a simple stochastic volatility model. In particular, we are evaluating the performance of such model on real data. We consider three indices S&P500, DOW JONES and STOXX50, in a time interval from Jan 1st, 2019 to June 30th, 2020, and from these, estimate the volatility computed from the model.

The indices have been selected as representatives of the global economic trends in the US and EU markets. Specifically, the S&P500 is a market-capitalization-weighted stock-price index tracing the performance of the 500 largest companies listed on US stock exchanges (NYSE and Nasdaq Exhange).
The DOW JONES, instead, is a price-weighted stock-market index and accounts for the 30 major companies listed on US stock exchanges, characterized for being "blue-chip". Also the EURO STOXX 50 follows bluce-chip stocks representing leading firms in regions of the Eurozone.\

The three plots below represent the time series of the log returns calculated for the three indices during our period of interest. In particular, each series is a sequence of daily observations representing the logarithm og the ratio between the closing price  of the index and the opening price of the index for a given day.

At a first visual inspection, we can see a common behaviour in the volatility of all three indices. Especially, it is worth pointing out the similarity between the pattern of DJI and that of S&P500 (especially in terms of peaks), which both differ from the STOXX50E, presumably because the first two describe the US stock market, while the last index describes the European one. For example, a difference that can be seen from a first glance at the plots is that the US indices log returns displays differences in fluctuation magnitude across different periods that are more marked than for the EU index, for which log returns display, in general, wider fluctuations. Anyways, what stands out the most in all three plots is the very wide fluctuations present from February - March 2020 to around September 2020, which, very intuitively, are connected with the Covid-19 Pandemic Crisis.

@@

Barndorff‐Nielsen, O. E., & Shephard, N. (2001). Non‐Gaussian Ornstein–Uhlenbeck‐based models and some of their uses in financial economics. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\

@@

Nakajima, J., & Omori, Y. (2012). Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student’s t-distribution. Computational Statistics & Data Analysis, 56(11), 3690-3704.\

>>>>

output:
  pdf_document: 
   latex_engine: xelatex
   fig_width: 5
   fig_height: 3.5
   fig_caption: true
   extra_dependencies: ["float"]
   number_sections: true 
---

\section{SV models + Literary Review}

In section (\textit{fill with the number of the section}) a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset. Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows, \begin{align*}
    y_t|x_t & \sim N(0,e^{x_t})\\
    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align*}

@@

Stochastic Volatility Models, whose early formulation is commonly attributed to Taylor (1982, 1986), allow to account for time-varying and autoregressive volatility in financial returns, posing themselves as a valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models in dealing with non-constant volatility. Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as \begin{align*}

@@

\end{align*} where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$. Kim, Shephard and Chib (1998) prefer $\psi=1$.} Most importantly, $\epsilon_t$ and $\eta_t$, the Gaussian white noise processes that drive the canonical model, are assumed to be uncorrelated. This latter assumption justifies another definition of this specification, namely "discrete SV model without leverage".\footnote{Note that in this review we do not focus on continuous time SV models. In fact, such models have attracted a considerable amount of research in financial econometrics and mathematical finance, especially after Hull and White (1987) considered stochastic volatility for option pricing. Arguably, the most influential model was then proposed for option pricing by Heston (1993), a SV model with leverage effects and square root diffusion driving volatility. Diffusion-based SV models enjoyed increasing popularity, see for example Barndorff-Nielsen and Shephard (2001), or Christoffersen et al. (2010), who investigate alternatives to the entrenched affine square root SV model.  Eraker (2004) proposed a SV model with correlated jumps in prices and volatility, extending Heston's model, while Comte and Renault (1998) extended Hull and White's model as to feature long memory properties. More recently, Gatheral et al. (2018) further built on Comte and Renault's fractional SV model to propose the popular "rough volatility models" (see also Friz et al., 2021).}\
In order to accommodate for leverage effects, namely capture the increase in volatility that follows a drop in the returns, the model can extended as in Omori et al. (2007), \begin{align*}

@@

\end{align*} where $\rho<0$ captures the negative correlation. Such specification, which can be referred to as "discrete SV model with leverage", captures the asymmetric response of volatility to returns of different signs, so that similar specifications are sometimes also deemed as "asymmetric SV models" (e.g. Harvey and Shephard 1996, Mao et al. 2020).\

@@

Assuming that $\epsilon_t$ follows a Student-$t$ distribution, and exploiting the fact that $\epsilon_t$ can then be written as $\lambda_t^{-1/2}\zeta_t$, where $\zeta_t \sim N(0,1)$ and $v\lambda_t \sim \chi^2_v$ (Harvey et al. 1994, Chib et al. 2002), we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004), \begin{align*}

@@

\end{align*} To capture other elements of the behaviour of financial data, several other extensions have been proposed. For example, SV models have been extended to include conditional heteroskedasticity in the mean returns (Koopman and Hol Uspensky 2002) to capture potential volatility feedback effects, or to feature autoregressive moving average innovations (Chan 2013, Zhang et al. 2020), allowing better goodness of fit and out-of-sample forecasts.\footnote{Dimitrakopoulos and Kolossiatis (2020) note that "the moving average component, the leverage effect and the conditional heteroscedasticity in mean have been considered separately in the stochastic volatility literature" and provide two specifications, one featuring an MA component and leverage effects, the other an MA component and conditional heteroskedasticity in mean.} \footnote{Other extensions, though ones for which we do not report the specifications, feature modelling the latent volatility process $x_t$ as an ARFIMA process (Long Memory Stochastic Volatility model, Breidt et al. 1998) or as governed by a first-order Markov process (Markov Switching Stochastic Volatility model, So et al. 1998). Recently, Luo et al. (2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as \begin{align*}

@@

\end{align*} while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads \begin{align*}

@@

\end{align*} where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\

@@

Finally, before moving to the application of Sequential Monte Carlo techniques to a SV model, we conclude by going through some references for the Bayesian analysis proposed for such models. Starting from the seminal work of Jacquier et al. (1994), the use of MCMC methods has become increasingly popular for parameter estimation and smoothing exercises in SV models (e.g. Kastner 2019, presenting the R package \textit{stochvol} for Bayesian parameter estimation, and Chopin and Papaspiliopoulos 2020, who use MCMC to sample from the smoothing distribution of a SV model). As regards filtering exercises, the adoption of particle filters was rather rapid:\footnote{Actually, SV models are now often used as straightforward applications of particle filters on non-linear state space models, see for example Andrieu et al. (2010), Douc et al. (2014) or Chopin and Papaspiliopoulos (2020).} indeed, latent volatilities in Kim et al. (1998) were already filtered by employing the particle filter suggested in Pitt and Shephard (1999), paving the way for subsequent applications (\textit{inter alia}) in Chib et al. (2006), Omori et al. (2007), Kim and Stoffer (2008) and Nakajima and Omori (2012).\

@@

As previously mentioned, we will analyze the behaviour of the described filtering tools associated to a simple stochastic volatility model. In particular, we are evaluating the performance of such model on real data. We consider the continuously compounded returns (also called logarithmic returns) of three indices, S&P500, DOW JONES and STOXX50, in a time interval from June 1st, 2017 to May 30th, 2021. From these, we estimate the daily volatility, as computed by the model.

The indices have been selected as representatives of the global economic trends in the US and EU markets. Specifically, the S&P500 is a market-capitalization-weighted stock-price index tracing the performance of the 500 largest companies listed on US stock exchanges (NYSE and Nasdaq Exchange). The DOW JONES, instead, is a price-weighted stock-market index and accounts for the 30 major companies listed on US stock exchanges, characterized for being "blue-chip". Also the EURO STOXX 50 follows bluce-chip stocks representing leading firms in regions of the Eurozone.\

The three plots below represent the time series of the log returns calculated for the three indices during our period of interest. In particular, each series is a sequence of daily observations representing the logarithm of the ratio between the closing price of the index for a given day and the closing price of the day before.

At a first visual inspection, we can see a common behaviour in the volatility of all three indices. Especially, it is worth pointing out the similarity between the pattern of DJI and that of S&P500 (especially in terms of peaks), which both differ from the STOXX50E, presumably because the first two describe the US stock market, while the last index describes the European one. For example, a difference that can be seen by having a glance at the plots is that the US indices log returns displays differences in fluctuation magnitude across different periods that are more marked than for the EU index, for which log returns display, in general, wider fluctuations. Anyways, what stands out the most in all three plots is the very wide fluctuations present from February - March 2020 to around September 2020, which, very intuitively, are connected with the Covid-19 Pandemic Crisis.

@@

Barndorff‐Nielsen, O. E., & Shephard, N. (2001). Non‐Gaussian Ornstein--Uhlenbeck‐based models and some of their uses in financial economics. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\

@@

Nakajima, J., & Omori, Y. (2012). Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student's t-distribution. Computational Statistics & Data Analysis, 56(11), 3690-3704.\
